Coalition Formation Algorithm of Prosumers in a Smart Grid Environment

### 2014-09-29 Article.tex
corrections

In this paper, we consider the case of prosumers grouping together into coalitions to be able to produce sufficient power which may be sold to the grid. If the coalition's production is sufficiently high and stable, the coalition can negociate on a day by day basis with the grid by announcing each day an expected production for the next day. We focus on how to statistically improve the production stability by carefully forming coalitions of prosumers that have greater probabilities of staying in acceptable range of production. By using prediction techniques, we form coalitions that meet the contract values ranges proposed by the grid, enabling it to schedule its production accordingly. Meeting contract values for the prosumers is especially relevant in power trade conditions, where energy is traded based on day-ahead predictions. In this context, participants should try to minimize their prediction errors in order to maintain a stable state for the grid. They may even occur some penalties if the productions deviate from the initial contract values. However, it remains pertinent for both side to maintain the production as stable as possible : for the prosumer, the stability (with renewable energy sources) will impact its net gain, and for the power grid, it will help to maintain its reliability.

@@

In this paper we do not consider stability against player defection but focus on a statistics-oriented definition of coalition stability. We ough indeed to form coalitions utilities that depend on statistical properties of time-series values. The setting is thus similar with some financial studies on stock exchanges, where researchers tried to find relevant correlated clusters of stocks based on their daily prices variations. There exists well-known algorithms such as K-means or hierarchical clustering that are traditionally used for such purposes. Nevertheless, despite their popularity, they do not appear particularly well suited for meaningfully translating complex correlation relationships between time series while clustering them, especially when these series exhibit complicated underlying patterns. In \cite{Mantegna1999} the author introduced an approach where stocks time-series of their daily log returns, are organized in a graph such that stocks exhibiting similar price fluctuation patterns are close to each other. This closeness notion is formalized with a similarity measure based on Pearson correlation coefficient ($ d_{ij} = \dfrac{1}{2}\sqrt{2(1-\rho_{ij})} $ or $ d_{ij} = 1 - \rho_{ij}^{2} $ ) that enables to weight the edges of the graph.

@@

Our major concern was to characterize a prosumer behavior with realistic patterns of consumption, production as well as realistic geographical correlations between them. Only wind turbines and PV will be used as generators throughout this paper, and we denote by $ \mathcal{A} = \{ a_{1},...a_{N} \} $ the set of prosumers. To account for the production of a prosumer, we relied on French's weather data from 2006 to 2012 sampled every 3 hours \cite{Infoclimat} (similar data are available for the United States covering 2010\cite{NCDC}). As shown in the first part of figure 1, the first step consists in discretizing the studied zone around well chosen weather stations and gathering three kinds of traces (wind speed, cloudiness, and temperature) that we organize in a structure we will refer to as "climate vector" in the following. We consider that climate vectors are constant over their area, meaning that, if two agents are in the same area, they are exposed to the same climate vector. We used existing DER models enabling conversion between weather variables and output powers (see power curves in \cite{Kota2011} \cite{windturbinemodel}). \cite{Dans2007} provides also a convenient way of using cloudiness traces as realistic degradation factors in a \textit{"clear blue sky"} solar model, which enables us to rebuild realistic solar irradiance traces for stations that do not provide such information directly.

@@

However, these clusters of strongly correlated time-series are the exact opposite of what we are seeking. We can indeed consider them directly as coalitions and compute their utilities, and the results show (see the green curves on figure \ref{Fig4}), as expected, terrible values (far worse than a random split of the agents in the same number of coalitions).

We thus opt for reversing the metric ($ \rho_{ij}^{2} $) such that uncorrelated time-series are close to each other in the graph and (anti)correlated time-series are distant. We defined a graph $G_\epsilon(\mathcal{A},E)$ where $\mathcal{A}$ is the set of agents and $ E $ the edge set. A given link between $ i $ and $ j $ is present if $ \rho_{ij}^{2} \leq \epsilon $, that is, $ e_{i,j} = \mathbf{1}_{\{\rho_{ij}^{2} \leq \epsilon \}}\rho_{ij}^{2} $. As expected \cite{Onnela2004}, independently of the $ \epsilon $ parameter, $ G_{\epsilon } $ exhibits henceforth much less clustering and communities seem hardly visible. Therefore, using classical clustering or community detection algorithms seem to provide poor results.

However, as seems intuitively understandable, cliques of this graph tend to exhibit very good utility values. Such  structures contain indeed a link between every two nodes, meaning that the overall correlation is quite small. Obviously, the $ \epsilon $ parameter is indirectly responsible for the sizes of the cliques : if it is too low, $ G_{\epsilon} $ does not provide enough cliques, conversely, if $\epsilon $ is too high, we loose important information as the graph becomes very dense and cliques tend to overlap strongly. For large values of $ \epsilon $ and for non trivial number of agents, finding cliques can even become computationally intensive. Despite being direct and simple, improving the sizes of the cliques by increasing $ \epsilon $ seems too brutal. Furthermore, as the utility function is also focused on maximizing $ P_{\phi}(s) $, it might be the case that some cliques benefit from additional agents even if they don't form a clique anymore.

Cliques for small values of $ \epsilon $ appear thus as good seeds for stable coalitions. We define $ N_{COAL} $ as the desired number of coalitions fixed by the user. Furthermore, let $ \Theta(G_{\epsilon}) $ be the set of non overlaping cliques in $ G_{\epsilon} $, we choose the $ \epsilon $ parameter as the smallest value such that $ G_{\epsilon} $ contains at least $ N_{COAL} $ non overlaping cliques :

@@

We construct $ G_{\epsilon^{\star}} $ such that it contains at least $ N_{COAL} $ potential seeds that may or may not be above the grid requirements, but exhibit low correlations among the members. Naturally, grid requirements could eventually be reached and social welfare increased by incorporating more nodes into the seeds. For this purpose, we use a clique percolation algorithm on $ G_{\epsilon^{\star}} $ that will extend the seeds over a correlation constrained environement. The main idea behind this algorithm is to make the seeds grow by local optimization of a fitness function, usually based on internal and external degrees of the seeds \cite{Lancichinetti}. The main difference between our case study and this algorithm for community detection is that we chose the fitness function as our utility function (cf. eq.\eqref{utility}) that both maximizes the production and minimizes its volatility.

As explained in section \ref{sec:related}, clique percolation leads generally to overlapping communities. For simplicity, in this paper, we wish to keep the coalitions disjoint and  leave the management of overlapping coalitions for future work. We thus implemented a simple heuristic that considers nodes in multiple seeds one by one and chooses its final coalition as the one that “needs it the most” in terms of utility loss. More formally, for a coalition s and a node $ i \in s $, we define :

@@

\begin{figure}
 \centering
  \includegraphics[scale=0.8]{./figure9/util}
\caption{Utility as function of the parameters space $\{\phi \in [0,1], N_{COAL}\in \mathbb{Z}^+, P^{MIN} \in \mathbb{R}^+\}$: a) In $ \{\phi, P^{MIN}\} $ when $ N_{COAL} $ is fixed. b) in $ \{N_{COAL}, P^{MIN}\} $ for a fixed $ \phi $. For readability, $ P^{MIN} $ is expressed in tenth of MW.}
\label{Fig2}
\end{figure}

@@

In figure \ref{Fig2}b, we fix the reliability to a given empirical value ($\phi = 0.08 $). We can observe how the social welfare $\sum_{s \in S} \mathcal{U}_{\phi,\ P^{MIN}}(s)$ evolves according to $P^{MIN}$ and $ N_{COAL} $. In our setting we notice a maximum point around $ N_{COAL} = 3 $, followed by a decrease. There is clearly a maximum number of coalitions sustainable for a given set of parameters. After this value we will form more coalitions that would be unable to pass the grid requirements. Moreover, reckon that increasing $ N_{COAL} $ means also increasing $ \epsilon $, leading to denser graphs, meaning that the algorithm performance will also decrease.

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.7]{./figure5/coals}
  \caption{Evolution of the coalitions for different values of $ P^{MIN} $ for clique percolation (blue dots) and random process (red dots). The diameters of the dots are proportional to the coalition sizes. The hatched zero utility zone corresponds to the \textit{"under-requirement space"}.}
  \label{Fig3}
\end{figure}

In figure \ref{Fig3}, we perform a simulation with 200 prosumers, and we fix the reliability $ \phi = 0.1 $ meaning that coalitions should produce more than their contract values at least $ 90 \% $ of the time. We observe the result of our algorithm for a fixed number of coalitions of $N_{COAL} = 10$ and let the algorithm form the coalitions as the grid changes according to $ P^{MIN} $. We compare our algorithm with a random process that splits the agents in $ N_{COAL} $ coalitions. Figure \ref{Fig3} shows this evolution for our algorithm (blue dots) and for the random process (red dots). The diameter of a dot is proportional to the number of agents in the coalition and the higher the dot is, the higher its utility. The $ P^{MIN} $ values of the x axis are expressed in tenth of MW for readability and the hatched zone corresponds to the \textit{"under-requirement space"}, meaning that whenever a coalition is in this zone, it has a null utility.

Looking at figure \ref{Fig3}, we can see first that our algorithm performs better than the random formation algorithm at finding high valued coalitions. The blue dots allowed to enter the market (non hatched zone) are indeed outnumbering the red dots, especially when the grid requirements are neither too low nor too high ($ P^{MIN} = 5 $ in figure \ref{Fig3}).

In more details, figure \ref{Fig4}a presents the evolution of social welfare as the number of coalitions increases (all other parameters are kept constant) for random (red curve), clique percolation (blue curve), and correlated coalitions (green curves) that stands for a worst case scenario. As for figure \ref{Fig4}b, it shows the percentage of coalitions able to enter the market for different values of $ P^{MIN} $. For consistency, we average the results of both plots over 100 realizations (colored zones around the curves shows the standard deviations).

When the grid requirements are constant (figure \ref{Fig4}b), and the number of desired coalitions is low, clique percolation generally performs better than random case. Moreover, when $ N_{COAL} $ gets bigger, the performance of a random split tumble down rapidly while our clique percolation is able to maintain efficiently the social welfare of the coalitions formed.

When the grid requirements vary, for very low $ P^{MIN} $, all coalitions for all algorithms are able to enter the market, yielding an acceptance percentage of $ 100 \% $. But as $ P^{MIN} $ increases, we see the percentage of the correlated coalitions is going down very quickly. After a few increases in $ P^{MIN} $, the percentage of the random procedure starts dropping while it stays almost constant for our algorithm. For $ P^{MIN} = 8 $, we see that only a little more than half of the coalitions for the random case are able to enter while approximately $ 85 \%$ of them enters for the clique percolation algorithm. Finally, when the grid requirements becomes too high, the acceptance percentage of our algorithm tends slowly to zero (not shown in this plot for readability).

@@

We believe that the originality of this paper lies in its willingness to exploit the de-correlation of prosumer profiles in order to build stable coalitions. In this direction, we presented a model based on meteorological traces, that captures the complex "\textit{energetic mix/climate vectors}" combination and generates realistic production and consumption patterns. We then built a framework that enables the grid to specify stability and minimum production requirements for filtering the coalitions. On this basis, we proposed a simple algorithm that seeks for uncorrelated prosumer patterns as potential seeds and expand them in coalitions able to rise above the grid requirements.

The algorithm is validated against a random process, where we show that it performs better (coalitions are more stable and the overall production is more important) and that it exhibits a higher robustness/flexibility against grid requirements changes.

>>>>

In this paper, we consider the case where prosumers wanted to be clustered together into coalitions in order to be able to produce enough power which may be sold to the grid. If the coalition's production is sufficiently high and stable, the coalition can negociate on a day by day basis with the grid by announcing each day an expected production for the next day. We focus on how to statistically improve the production stability by carefully forming coalitions of prosumers that have greater probabilities of staying in acceptable range of production. By using prediction techniques, we form coalitions that meet the contract values ranges proposed by the grid, enabling it to schedule its production accordingly. Meeting contract values for the prosumers is especially relevant in power trade conditions, where energy is traded based on day-ahead predictions. In this context, participants should try to minimize their prediction errors in order to maintain a stable state for the grid. They may even occur some penalties if the productions deviate from the initial contract values. However, it remains pertinent for both side to maintain the production as stable as possible : for the prosumer, the stability (with renewable energy sources) will impact its net gain, and for the power grid, it will help to maintain its reliability.

@@

In this paper we do not consider stability against player defection but focus on a statistics-oriented definition of coalition stability. We want to form coalitions based on utilities that depend on statistical properties of time-series values. The setting is thus similar with some financial studies on stock exchanges, where researchers tried to find relevant correlated clusters of stocks based on their daily prices variations. There exists well-known algorithms such as K-means or hierarchical clustering that are traditionally used for such purposes. Nevertheless, despite their popularity, they do not appear particularly well suited for meaningfully translating complex correlation relationships between time series while clustering them, especially when these series exhibit complicated underlying patterns. In \cite{Mantegna1999} the author introduced an approach where stocks time-series of their daily log returns, are organized in a graph such that stocks exhibiting similar price fluctuation patterns are close to each other. This closeness notion is formalized with a similarity measure based on Pearson correlation coefficient ($ d_{ij} = \dfrac{1}{2}\sqrt{2(1-\rho_{ij})} $ or $ d_{ij} = 1 - \rho_{ij}^{2} $ ) that enables to weight the edges of the graph.

@@

Our major concern is to characterize a prosumer behavior with realistic patterns of consumption, production as well as realistic geographical correlations between them. Only wind turbines and PV will be used as generators throughout this paper, and we denote by $ \mathcal{A} = \{ a_{1},...a_{N} \} $ the set of prosumers. To account for the production of a prosumer, we rely on French's weather data from 2006 to 2012 sampled every 3 hours \cite{Infoclimat} (similar data are available for the United States covering 2010\cite{NCDC}). As shown in the first part of figure 1, the first step consists in discretizing the studied zone around well chosen weather stations and gathering three kinds of traces (wind speed, cloudiness, and temperature) that we organize in a structure we will refer to as "climate vector" in the following. We consider that climate vectors are constant over their area, meaning that, if two agents are in the same area, they are exposed to the same climate vector. We use existing DER models enabling conversion between weather variables and output powers (see power curves in \cite{Kota2011} \cite{windturbinemodel}). \cite{Dans2007} provides also a convenient way of using cloudiness traces as realistic degradation factors in a \textit{"clear blue sky"} solar model, which enables us to rebuild realistic solar irradiance traces for stations that do not provide such information directly.

@@

However, these clusters of strongly correlated time-series are the exact opposite of what we are seeking. We can indeed consider them directly as coalitions and compute their utilities, and the results show (see the green curves on figure \ref{Fig4}), as expected, terrible values (far worse than a random split of the agents in the same number of coalitions). We thus opt for reversing the metric ($ \rho_{ij}^{2} $) such that uncorrelated time-series are close to each other in the graph and correlated time-series are distant. We defined a graph $G_\epsilon(\mathcal{A},E)$ where $\mathcal{A}$ is the set of agents and $ E $ the edge set. A given link between $ i $ and $ j $ is present if $ \rho_{ij}^{2} \leq \epsilon $, that is, $ e_{i,j} = \mathbf{1}_{\{\rho_{ij}^{2} \leq \epsilon \}}\rho_{ij}^{2} $. As expected \cite{Onnela2004}, independently of the $ \epsilon $ parameter, $ G_{\epsilon } $ exhibits henceforth much less clustering and communities seem hardly visible. Therefore, using classical clustering or community detection algorithms seem to provide poor results.

However, as seems intuitively understandable, cliques of this graph tend to exhibit very good utility values. Such  structures contain indeed a link between every two nodes, meaning that the overall correlation is quite small. Obviously, the $ \epsilon $ parameter is indirectly responsible for the sizes of the cliques : if it is too low, $ G_{\epsilon} $ does not provide enough cliques, conversely, if $\epsilon $ is too high, we loose important information as the graph becomes very dense and cliques tend to overlap strongly. For large values of $ \epsilon $ and for non trivial number of agents, finding cliques can even become computationally intensive. Despite being direct and simple, improving the sizes of the cliques by increasing $ \epsilon $ seems too brutal. Furthermore, as the utility function is also focused on maximizing $ P_{\phi}(s) $, it might be the case that some cliques benefit from additional agents even if they don't form a clique anymore. Cliques for small values of $ \epsilon $ appear thus as good seeds for stable coalitions. We define $ N_{COAL} $ as the desired number of coalitions fixed by the user. Furthermore, let $ \Theta(G_{\epsilon}) $ be the set of non overlaping cliques in $ G_{\epsilon} $, we choose the $ \epsilon $ parameter as the smallest value such that $ G_{\epsilon} $ contains at least $ N_{COAL} $ non overlaping cliques :

@@

We construct $ G_{\epsilon^{\star}} $ such that it contains at least $ N_{COAL} $ potential seeds that may or may not be above the grid requirements, but exhibit low correlations among the members. Naturally, grid requirements could eventually be reached and social welfare increased by incorporating more nodes into the seeds. For this purpose, we use a clique percolation algorithm on $ G_{\epsilon^{\star}} $ that will extend the seeds over a correlation constrained environment. The main idea behind this algorithm is to make the seeds grow by local optimization of a fitness function, usually based on internal and external degrees of the seeds \cite{Lancichinetti}. The main difference between our case study and this algorithm for community detection is that we chose the fitness function as our utility function (cf. eq.\eqref{utility}) that both maximizes the production and minimizes its volatility. As explained in section \ref{sec:related}, clique percolation leads generally to overlapping communities. For simplicity, in this paper, we wish to keep the coalitions disjoint and  leave the management of overlapping coalitions for future work. We thus implemented a simple heuristic that considers nodes in multiple seeds one by one and chooses its final coalition as the one that “needs it the most” in terms of utility loss. More formally, for a coalition s and a node $ i \in s $, we define :

@@

\begin{figure}
 \centering
  \includegraphics[scale=0.75]{./figure9/util}
\caption{Utility as function of the parameters space $\{\phi \in [0,1], N_{COAL}\in \mathbb{Z}^+, P^{MIN} \in \mathbb{R}^+\}$: a) In $ \{\phi, P^{MIN}\} $ when $ N_{COAL} $ is fixed. b) in $ \{N_{COAL}, P^{MIN}\} $ for a fixed $ \phi $. For readability, $ P^{MIN} $ is expressed in tenth of MW.}
\label{Fig2}
\end{figure}

@@

In figure \ref{Fig2}b, we fix the reliability to a given empirical value ($\phi = 0.08 $). We can observe how the social welfare $\sum_{s \in S} \mathcal{U}_{\phi,\ P^{MIN}}(s)$ evolves according to $P^{MIN}$ and $ N_{COAL} $. In our setting we notice a maximum point around $ N_{COAL} = 3 $, followed by a decrease. There is clearly a maximum number of coalitions sustainable for a given set of parameters. After that value the coalitions newly formed won't be able to pass the grid requirements. Moreover, reckon that increasing $ N_{COAL} $ means also increasing $ \epsilon $, leading to denser graphs, meaning that the algorithm performance will also decrease.

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{./figure5/coals}
  \caption{Evolution of the coalitions for different values of $ P^{MIN} $ for clique percolation (blue dots) and random process (red dots). The diameters of the dots are proportional to the coalition sizes. The hatched zero utility zone corresponds to the \textit{"under-requirement space"}.}
  \label{Fig3}
\end{figure}

In fig. \ref{Fig3}, we perform a simulation with 200 prosumers, and we fix the reliability $ \phi = 0.1 $ meaning that coalitions should produce more than their contract values at least $ 90 \% $ of the time. We observe the result of our algorithm for a fixed number of coalitions of $N_{COAL} = 10$ and let the algorithm form the coalitions as the grid changes according to $ P^{MIN} $. We compare our algorithm with a random process that splits the agents in $ N_{COAL} $ coalitions. Fig. \ref{Fig3} shows this evolution for our algorithm (blue dots) and for the random process (red dots). The diameter of a dot is proportional to the number of agents in the coalition and the higher the dot, the higher its utility. The $ P^{MIN} $ values of the x axis are expressed in tenth of MW for readability and the hatched zone corresponds to the \textit{"under-requirement space"}, meaning that whenever a coalition is in this zone, it has a null utility. Looking at fig. \ref{Fig3}, we can see first that our algorithm performs better than the random formation algorithm at finding high valued coalitions. The blue dots allowed to enter the market (non hatched zone) are indeed outnumbering the red dots, especially when the grid requirements are neither too low nor too high ($ P^{MIN} = 5 $ in fig. \ref{Fig3}).

In more details, figure \ref{Fig4}a presents the evolution of social welfare as the number of coalitions increases (all other parameters are kept constant) for random (red curve), clique percolation (blue curve), and correlated coalitions (green curves) that stands for a worst case scenario. As for figure \ref{Fig4}b, it shows the percentage of coalitions able to enter the market for different values of $ P^{MIN} $. For consistency, we average the results of both plots over 100 realizations (colored zones around the curves shows the standard deviations). When the grid requirements are constant (figure \ref{Fig4}b), and the number of desired coalitions is low, clique percolation generally performs better than random case. Moreover, when $ N_{COAL} $ gets bigger, the performance of a random split tumble down rapidly while our clique percolation is able to maintain efficiently the social welfare of the coalitions formed. When the grid requirements vary, for very low $ P^{MIN} $, all coalitions for all algorithms are able to enter the market, yielding an acceptance percentage of $ 100 \% $. But as $ P^{MIN} $ increases, we see the percentage of the correlated coalitions is going down very quickly. After a few increases in $ P^{MIN} $, the percentage of the random procedure starts dropping while it stays almost constant for our algorithm. For $ P^{MIN} = 8 $, we see that only a little more than half of the coalitions for the random case are able to enter while approximately $ 85 \%$ of them enters for the clique percolation algorithm. Finally, when the grid requirements becomes too high, the acceptance percentage of our algorithm tends slowly to zero (not shown in this plot for readability).

@@

We believe that the originality of this paper lies in its willingness to exploit the de-correlation of prosumer profiles in order to build stable coalitions. In this direction, we presented a model based on meteorological traces, that captures the complex "\textit{energetic mix/climate vectors}" combination and generates realistic production and consumption patterns. We then built a framework that enables the grid to specify stability and minimum production requirements for filtering the coalitions. On this basis, we proposed a simple algorithm that seeks for uncorrelated prosumer patterns as potential seeds and expand them in coalitions able to rise above the grid requirements. We validated our algorithm against a random choice of coalitions, which can be considered as an average case. We showed  that it performs better (coalitions are more stable and the overall production is more important) and that it exhibits a higher robustness/flexibility against grid requirements changes. The worst case scenario is represented by coalitions of correlated prosumers.
